Formal Grammars as Types in Non-Commutative Linear-Non-Linear Type Theory

### 2024-07-10 POPL/paper.tex
gonna add some future/related work then sleep

\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\Sigma^{*}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\true}{\mathbf{true}}
\newcommand{\false}{\mathbf{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Grammar Type Theory\xspace}
\newcommand{\theoryabbv}{GTT\xspace}

@@

The theory of formal languages and parsing is one of the oldest and
most thoroughly developed areas of theoretical computer science. A
prominent topic in the 1950, 60s, and 70s led to a series of remarkable
developments: Chomsky's hierarchy of grammar formalisms \cite{chomThreeModels1956}, practical
algorithms for parsing of regular and context-free grammars and
variants \cite{KNUTH1965607,Earley1970}, as well as implementations of practical tools for the
generation of efficient parsers \cite{Johnsonyacc}.

@@

@@

Inside of an LNL system, linearity may be thought of as an option that users can
choose to deliberately invoke at deliberate points in their developments in an
otherwise intuitionistic system. However, if we are wishing to treat parsers as
linear terms over input strings, the non-linear fragment of an LNL theory does
not really assist in the development of parsers. It is instead the case that
parsers may benefit from a \emph{dependence} on non-linear terms.
Through the approach described by Krishaswami et al.\ in
\cite{krishnaswami_integrating_2015},
we may define a restricted form of dependent types. In particular, dependence
on linear terms
is disallowed; however, through dependence of a linear term on a non-linear
index, we recover the definition of Aho's indexed grammars \cite{AhoIndexed}
internal to \theoryabbv.

@@

\paragraph{Linear Function Types}
The monoidal structure provided by $\otimes$ is both left and right closed and
this is denoted by the left and right linear function types.

\steven{Say more about the adjunctions}

Further, for linear types $A$ and $B$ the left linear function type $A \lto B$
enjoys an elimination principle similar to function application/modus ponens.

@@

That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
complete to parses of $B$. In this manner, the linear function types generalize
Brzozowksi's notion of derivative
\cite{brzozowskiDerivativesRegularExpressions1964}.
Brzozowski initially only gave an accounting of this operation for
generalized regular expressions, but later work from Might et al.\ demonstrates that the same
construction can be generalized to context free grammars
\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
types the same notion of derivative extends to the grammars of \theoryabbv.

@@

@@

\paragraph{Empty Type}
The empty type $\bot$ has no inhabitants. The elimination for the empty type
witnesses the principle of explosion --- i.e. from a term of type $\bot$ we may
introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.

@@

\paragraph{The $G^{-1}$ Modality}
Turn nonlinear into linear

\steven{Call this $F$ or $G^{-1}$?}

\steven{Why can't these map any nonlinear type into linear ones? instead of just
those that are already linear types lifted by $G$?}

\steven{TODO G inverse}
  % \item $\sem{G A} \gamma = \sem{A} \gamma \varepsilon$

@@

\subsection{Subtheories}
\paragraph{Regular Expressions} We may realize regular expression as a subtheory
of our language by restricting the type constructors to the
linear unit, characters, $\otimes$, $\oplus$, and Kleene star. Classically, the
languages recognized by these are exactly the regular languages --- the lowest
on the Chomksy hierarchy.

\paragraph{$\mu$-Regular Expressions} Similarly, we may restrict the connectives
to be linear unit, characters, $\otimes$, $\oplus$, and arbitrary recursion.
Instead of regular expressions these correspond to Lei{\ss}'s $\mu$-regular
expressions, which are known to be equivalent to context-free grammars
\cite{leiss}.

\steven{Give this subtheory a name }

@@

\steven{TODO $\mu$ regular expressions}

@@

\section{Nonstandard Models}
\label{sec:nonstandard}

@@

\subsection{Parse Canonicity}
Canonicity is an important metatheoretic theorem in the type theory
literature.  It provides insight on the normal forms of terms and,
therefore, on its computational aspects. Frequently, proving
canonicity for boolean types, i.e. every closed term of type bool
reduces to either true or false, is enough to justify that the type
theory being studied is well-behaved. In our case, however, since we
want to connect \theoryabbv to parsers, we must provide a more
detailed account of canonicity.

@@

\section{Automata as Grammars}
\label{sec:automata}
\pedro{The switch from the canonicity proof to this section feels a bit ``bumpy''.
  Since they are at least technically related, it might be a good idea to say something
  like ``The canonicity theorem from the previous section gives a tight connection
  between terms in our type theory and the theory of formal languages. In this section
we further explore these connections by using the fact that, clasically ...''}

Classically, formal language theory is closely related to the study of automata.
In the Chomsky hierarchy, each language class is associated to a class of
automata that serve as recognizers. Internal to \theoryabbv,
we can characterize these language classes syntactically; moreover, we
demonstrate the equivalence of these language classes to its associated automata class as a
proof term within our logic.

Internalizing these equivalences is particularly useful for building a regular
expression parser. That is, it is relatively easy for use to write down a parser
for deterministic finite automata (DFAs). Then we may compose this parser with
a proof equivalence between DFAs and regular expressions to extract a verified
regular expression parser.

As we will see in \cref{subsubsec:tm}, we may very quickly encode arbitrary
recursively enumerable languages --- since we can simulate the
tape of Turing machine in the dependent fragment of the language.We reflect
here on this expressive power to motivate why we put stringent syntactic
restrictions on particular classes of automata. The type theory we present truly
is general enough to easily capture all of the Chomsky hierarchy.
To begin, let's focus on the simplest languages and their corresponding
automata --- regular expressions and finite automata.
\pedro{The connection dependent types-Turing machines seems too important to only
  be mentioned here. It should at least indirectly motivate the type theory being
dependently typed.}

@@

Meaning, each state has a transition for every character.

@@

% \todoin{Derivation tree for this construction}

Perhaps this derivation is not too surprising. All it says
is that a DFA may be ran by transitioning a single character
at a time, and then accepting or rejecting based on the
final state. This is exactly what DFAs did initially, so
what did we gain? Well, this has the benefit of our type
system to ensure its correctness. Moreover, this construction exports to an
intrinsically verified and executable DFA via Agda.

\subsection{Regular Grammars and DFAs}
\label{subsec:deriv}

In order to extend the DFA
parser from \cref{subsec:regexparser} to the construction of
a verified parser
generator for regular grammars we need to perform some
plumbing establishing an equivalence between regular
expressions and DFAs.

@@

\begin{theorem}[Thompson]
  \label{thm:thompson}
  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
  language as $g$.
\end{theorem}
\steven{In Agda as a weak equivalence right now}

We make a pretty straightforward adaptation of Thompson's theorem to our setting,

@@

\begin{proof}[Proof Sketch]
  The following proof is currently formalized in Agda. Below
  we give a walk through the structure of the proof.

  Recall that regular grammars are inductively defined via
  disjunction, concatenation, and Kleene star over literals
  and the empty grammar. By induction over regular grammars,
  we will construct an NFA that is equivalent to $g$.

@@

  From here, to prove
  $I \cong \mathsf{AccTrace}_{NFA(I)}$ it suffices to show
  $I \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$. Below we
  give two parse transformers, one from $I$ to
  $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa. The formalized
  artifact provides the low-level details that these are
  indeed mutually inverse.

@@

  \cref{fig:disjunctionNFA} shows the process of
  disjunctively combining NFAs. Precisely, we add a single
  new state and we included copies of the states from each of $N$
  and $M$. The new state acts as the initial state and has
  $\varepsilon$-transitions to the initial states of $N$ and
  $M$. We include all of the internal transitions from $N$
  and $M$, and the accepting states of $N \oplus_{NFA} M$
  are exactly the accepting states in each subautomaton.

@@

  As discussed above, the other cases and low-level details
  of isomorphism are reserved for the formalization. This
  concludes the proof of our variant of Thompson's construction.
\end{proof}

@@

We encode the traces of a pushdown automaton very similarly to those of an NFA,
except the transitions of a PDA are instead encoded via the linear-non-linear
$\Pi$-type. This is because of simply transitioning via character, a PDA must
also pop and push characters onto a stack, which is used as the argument to
these dependent functions.

@@

In
\cref{subsubsec:pda}, we gave a grammar presentation of
traces through a PDA by using a non-linear type $S$
to encode the stack. We may similarly use pairs $S \times S$
to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
infinite tape of a Turing machine. The intuition behind this correspondence is
that the left half of the tape is on one stack, the right
half of the tape the other, and we treat the tops of stacks
like the head of the tape.

Due to computability limitations, we are not able to extract
a verified parser for all recursively enumerable languages.
Yet we should still reflect on the generality of our syntax,
as it allows a natural encoding of Turing machines.

>>>>

\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\Sigma^{*}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathbf{Bool}}
\newcommand{\true}{\mathbf{true}}
\newcommand{\false}{\mathbf{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Grammar Theoretic Type Theory\xspace}
\newcommand{\theoryabbv}{GTTT\xspace}

@@

The theory of formal languages and parsing is one of the oldest and
most thoroughly developed areas of theoretical computer science. A
prominent topic in the 1950, 60s, and 70s led to a series of remarkable
developments: Chomsky's hierarchy of grammars \cite{chomThreeModels1956}, practical
algorithms for parsing of regular and context-free grammars and
variants \cite{KNUTH1965607,Earley1970}, as well as implementations of practical tools for the
generation of efficient parsers \cite{Johnsonyacc}.

\steven{I don't like the name \theoryname (\theoryabbv). Feel free to play with it though, as
  its set up as a pair of macros}

@@

\steven{I initially wanted to say ``universal semantic domain to reason about
  parsing'', but these words have specific domain theory usage that aren't
  appropriate here, right?}

@@

Inside of an LNL system, linearity may be thought of as an option that users can
choose to deliberately invoke at deliberate points in their developments in an
otherwise intuitionistic system. However, if we are wishing to treat parsers as
linear terms over input strings, the non-linear fragment of an LNL theory does
not really assist in the development of parsers. It is instead the case that
parsers may benefit from a \emph{dependence} on non-linear terms.
Through the approach described by Krishnaswami et al.\ in
\cite{krishnaswami_integrating_2015},
we may define a restricted form of dependent types. In particular, dependence
on linear terms
is disallowed; however, through dependence of a linear term on a non-linear
index, we recover the definition of Aho's indexed grammars \cite{AhoIndexed}
internal to \theoryabbv.

\steven{The above is meant to walk the reader into accepting LNL as a natural
  choice, but it doesn't really flow well to the beginning of the next subsection }

@@

\paragraph{Linear Function Types}
The monoidal structure provided by $\otimes$ is both left and right closed and
this is denoted by the left and right linear function types. Further,
for linear types $A$ and $B$ the left linear function type $A \lto B$
enjoys an elimination principle similar to function application or modus ponens.

@@

That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
complete to parses of $B$. In this manner, the linear function types generalize
Brzozowksi's notion of derivative
\cite{brzozowskiDerivativesRegularExpressions1964}.
Brzozowski initially only gave an accounting of this operation for
generalized regular expressions, but later work from Might et al.\ demonstrates that the same
construction can be generalized to context free grammars
\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
types, the same notion of derivative extends to the grammars of \theoryabbv.
Note, to ensure that the linear function types do indeed generalize Brzozowski
derivative, we must include the equalities in \cref{fig:brzozowskideriv} as axioms.

@@

\begin{figure}
\begin{align*}
  c\lto c &\equiv I\\
  c\lto d &\equiv \bot\\
  c\lto I &\equiv \bot\\
  c\lto 0 &\equiv \bot\\
  c\lto (A \otimes B) & \equiv (c\lto A) \otimes B + (A \amp I) \otimes (c\lto B)\\
  c\lto A^* &\equiv (A \amp I)^{*} \otimes (c \lto A) \otimes A^* \\
  c\lto (A + B) &\equiv (c\lto A) + (c\lto B)
\end{align*}
\caption{Equality for Brzozowski Derivatives}
\label{fig:brzozowskideriv}
\end{figure}

@@

\paragraph{Empty Type}
The empty type $\bot$ has no inhabitants. The elimination for the empty type
witnesses the principle of explosion --- i.e.\ from a term of type $\bot$ we may
introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.

@@

\paragraph{The $G$ Modality}
The $G$ modality provides the embedding of linear types in the empty context
into non-linear types. The introduction and elimination forms for $G A$ obey the
same rules as given in \cite{bentonMixedLinearNonlinear1995} and \cite{krishnaswami_integrating_2015}.

The left adjoint $F$ from non-linear types back to linear types may be defined
as,

\[
  F X := \LinSigTy a A I
\]

\steven{Elaborate on $G$ more}

The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.

\[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]

@@

\subsection{Subtheories}
Each choice of connective has direct implications on the expressivity of
\theoryabbv. In particular, the set of connectives used will determine where the
grammar semantics is placed in
the Chomsky hierarchy.

\paragraph{Regular Expressions} We may realize regular expression as a subtheory
of our language by restricting the type constructors to the
linear unit, characters, $\otimes$, $\oplus$, and Kleene star. Classically, the
languages recognized by these are exactly the regular languages --- the lowest
on the Chomksy hierarchy.

\paragraph{$\mu$-Regular Expressions} Similarly, we may restrict the connectives
to be linear unit, characters, $\otimes$, $\oplus$, and arbitrary recursion via
left-fixed point rather than only Kleene star.
Instead of regular expressions these correspond to Lei{\ss}'s $\mu$-regular
expressions, which are known to be equivalent to context-free grammars
\cite{leiss,krishnaswami_typed_2019}.

@@

@@

\section{Other Models}
\label{sec:othermodels}
\subsection{Language Semantics}
Every grammar induces a language semantics. Also languages can be taken as a
propositionally truncated view of the syntax. Logical equivalence should induce
weak equivalence, and thus even give a syntactic way to reason about language equivalence.
\steven{TODO language semantics}

@@

\subsection{Parse Canonicity}
Canonicity is an important metatheoretic theorem in the type theory
literature.  It provides insight on the normal forms of terms and,
therefore, on its computational aspects. Frequently, proving
canonicity for boolean types, i.e. every closed term of type bool
reduces to either true or false, is enough to justify that the type
theory being studied is well-behaved. In our case, however, since we
want to connect \theoryabbv to parsers, we must provide a more
detailed account of canonicity. In particular, we give a nonstandard semantics
of \typeabbv that carries a proof of canonicity along with it.

@@

\section{Automata as Grammars}
\label{sec:automata}

The canonicity theorem from the previous section gives a tight connection
between terms in our type theory and the theory of formal languages. In this section
we further explore these connections by using the fact that, clasically
formal language theory is closely related to the study of automata.
In the Chomsky hierarchy, each language class is associated to a class of
automata that serve as recognizers. Internal to \theoryabbv,
we can characterize these language classes syntactically; moreover, we
demonstrate the equivalence of these language classes to its associated automata class as a
proof term within our logic.

@@

Meaning, each state has a transition for every character. \emph{Note:} the above use
  of a disjunction over $\Sigma$ is a bit of abuse of notation.
  As $\Sigma$ is a finite alphabet, we wish to think of this as an iterated
  binary sum, or perhaps as a finitely indexed sum defined via linear dependent
  sums in the same manner by which we defined binary sums.

@@

Perhaps this derivation is not too surprising. All it says
is that a DFA may be ran by transitioning a single character
at a time, and then accepting or rejecting based on the
final state. This is exactly what DFAs did initially, so
what did we gain? Well, this has the benefit of our type
system to ensure its correctness. Moreover, this construction exports to an
intrinsically verified and executable decision procedure for DFAs in \thoeryabbv
embedded in Agda.

We should note that the
construction of a DFA parser requires the addition of an additional but seemingly innocuous
axiom.

\[
  \top \cong \left( \bigoplus_{c : \Sigma} c \right)^{*}
\]

That is, the characters in the alphabet $\Sigma$ really are all of the
characters. The necessity of this axiom arises when we go to take a transition,
as we must ensure that the nexts character read has a corresponding transition
--- which would be guaranteed by determinism provided there are no surprise characters.
\steven{The above paragraph sounds weird}

\subsection{Regular Expressions and DFAs}
\label{subsec:deriv}

In order to extend the DFA
parser from \cref{subsec:regexparser} to the construction of
a verified parser
generator for regular expression we need to perform some
plumbing establishing an equivalence between regular
expressions and DFAs.

@@

\begin{theorem}[Thompson]
  \label{thm:thompson}
  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same
  language as $g$.
\end{theorem}
We make a pretty straightforward adaptation of Thompson's theorem to our setting,

@@

\begin{proof}[Proof Sketch]
  Recall that regular grammars are inductively defined via
  disjunction, concatenation, and Kleene star over literals
  and the empty grammar. By induction over regular grammars,
  we will construct an NFA that is equivalent to $g$.

@@

  From here, to prove
  $I \cong \mathsf{AccTrace}_{NFA(I)}$ it suffices to show
  $I \cong \mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$. Below we
  give two parse transformers, one from $I$ to
  $\mathsf{Trace}_{NFA(I)}(q_{1} , q_{2})$ and vice versa.

@@

  \Cref{fig:disjunctionNFA} shows the process of
  disjunctively combining NFAs. Precisely, we add a single
  new state and we included copies of the states from each of $N$
  and $M$. The new state acts as the initial state and has
  $\varepsilon$-transitions to the initial states of $N$ and
  $M$. We include all of the internal transitions from $N$
  and $M$, and the accepting states of $N \oplus_{NFA} M$
  are exactly the accepting states in each subautomaton.

@@

  When you wish to show that these actually maps invert each other to form an isomorphism, we invoke the
  universal properties of the $\mathsf{Trace}$ types. That is,
  $\mathsf{Trace}_{\oplus NFA}$, $\mathsf{Trace}_{N}$, and $\mathsf{Trace}_{M}$
  are \emph{initial} algebras to the the functors that define them. Thus, the
  embedding of traces through $N$ into $\oplus NFA$ and then mapped back down to $N$
  compose to a map from $\mathsf{Trace}_{N}$ to itself. Such a map is unique and
  thus must be the identity. The same reasoning shows that the desired
  composite do indeed invert each other to form an isomorphism.

  As discussed above, the other cases and
  follow through a similar argument. This
  concludes the proof of our variant of Thompson's construction.
\end{proof}

@@

We encode the traces of a pushdown automaton very similarly to those of an NFA,
except the transitions of a PDA are instead encoded via the LNL product type
$\bigamp$. This is because when simply transitioning via character, a PDA must
also pop and push characters onto a stack, which is used as the argument to
these dependent functions. Thus, the appropriate formation of traces through a
PDA is dependent on a stack.

@@

Above we gave a grammar presentation of
traces through a PDA by using a non-linear type $S$
to encode the stack. We may similarly use pairs $S \times S$
to encode the tape of a Turing machine. With two stacks we can simulate the behavior of the
infinite tape of a Turing machine. The intuition behind this correspondence is
that the left half of the tape is on one stack, the right
half of the tape the other, and we treat the tops of stacks
like the head of the tape. We should reflect on the generality of our syntax, as
it allows a natural encoding of Turing machines.
